Dependent Types for Safe and Secure Web Programming

### 2013-11-15 paper/protocols.tex
Paper revisions

In \idris{}, types are first-class, meaning that they may be treated like other
terms in computations. We may therefore define the following type synonym,
which we may use in order to make use of the CGI framework:

@@

SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database
engine often used as simple, structured storage for larger applications. We
make use of SQLite due to its simplicity, although we envisage that these
concepts would be applicable to more complex database management systems.

@@

Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
access, which we have implemented for the SQLite library. This is encapsulated
by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
bind and retrieve data types other than \texttt{String}, in the interest of
brevity).

\begin{SaveVerbatim}{databaseeff}
{-                           { Input resource type }    { Output resource type }             { Value }           -}
data Sqlite : Effect where
  OpenDB            : DBName -> 
                      Sqlite ()                         (Either () SQLiteConnected)          (Either SQLiteCode ())
  CloseDB           : Sqlite (SQLiteConnected)          ()                                   ()
  PrepareStatement  : QueryString -> 
                      Sqlite (SQLiteConnected)          (Either (SQLitePSFail) 
                                                          (SQLitePSSuccess Binding))         (Either SQLiteCode ())
  BindInt           : ArgPos -> Int -> 
                      Sqlite (SQLitePSSuccess Binding)  (SQLitePSSuccess Binding)            ()
  FinishBind        : Sqlite (SQLitePSSuccess Binding)  (Either SQLiteFinishBindFail
                                                          (SQLitePSSuccess Bound))           (Either BindError ())
  ExecuteStatement  : Sqlite (SQLitePSSuccess Bound)    (Either (SQLiteExecuting InvalidRow)
                                                          (SQLiteExecuting ValidRow)         StepResult
  RowStep           : Sqlite (SQLiteExecuting ValidRow) (Either (SQLiteExecuting InvalidRow)
                                                          (SQLiteExecuting ValidRow))        StepResult
  GetColumnText     : Column -> 
                      Sqlite (SQLiteExecuting ValidRow) (SQLiteExecuting ValidRow)           String

@@

%TODO: Fix up the labels
\begin{figure}[h]
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, node distance=6em, font={\sffamily\bfseries}]

@@

  \node[initial,state] (A)              {Uninitialised};
  \node[state]         (B) [below of=A] {ConnectionOpened};
  \node[state]         (C) [below of=B] {PreparedStatementOpen};
  \node[state]         (D) [below of=C] {PreparedStatementBinding};
  \node[state]         (E) [below of=D] {PreparedStatementBound};
  \node[state]         (F) [below of=E] {PreparedStatementExec};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{openDB}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=3em] {\texttt{closeDB, connFail}} (A)
        (B) edge[stateEdge]   node[edgeLabel, xshift=4em] {\texttt{prepareStatement}} (C)
        (C) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=3em, yshift=-4em] {\texttt{stmtFail}} (A)
        (C) edge[stateEdge]   node[edgeLabel, xshift=3em] {\texttt{startBind}} (D)
        (D) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{finishBind}} (E)
        (E) edge[stateEdge, bend left=85]   node[edgeLabel, xshift=-3.5em] {\texttt{bindFail}} (A)
        (E) edge[stateEdge]   node[edgeLabel, xshift=3.5em] {\texttt{beginExecution}} (F)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=-3.5em] {\texttt{finalise}} (B)
        (F) edge[stateEdge, bend right=85]   node[edgeLabel, xshift=3.5em] {\texttt{executeFail}} (A)
            edge[stateEdge, loop left]   node[edgeLabel, text width=2cm, xshift=2em] {\texttt{nextRow, reset}} (B);
\end{tikzpicture}
}
\caption{Database Resource Usage Protocol}
\label{fig:sqlitestates}
\end{figure}

@@

\end{SaveVerbatim}
\useverb{sqliteexecuting}

Column access functions may only be called when there is a valid row in the buffer, as signified by the input resource of each operation.

@@

\noindent
In order to provide the necessary static guarantee that there is a valid row to process in the buffer, we make use of the \texttt{if\_valid} construct. The \texttt{nextRow} function 
%If a failure happens at any point during the computation, the resource is
%updated to reflect the failure: if, for example, the library failed to create a
%connection to the database, the resource value would be updated to
%\texttt{OpenConn InvalidConn}. At this point, no further side-effecting
%requests are made to the underlying SQLite library, in order to ensure safety.
%The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and
%\texttt{executeFail} utility functions allow for failures, once detected, to be
%handled by executing the appropriate sequence of state transition functions to
%dispose of any open resources and return to the initial protocol state.

@@

\noindent
Each call to \texttt{nextRow}, which is a wrapper around the underlying
\texttt{sqlite3\_step()} C library function, returns a result of type
\texttt{StepResult}, which is then reflected in the resource. Calls to
\texttt{sqlite3\_step()} are only executed if the previous \texttt{StepResult}
is either \texttt{Unstarted} or \texttt{StepComplete}. We may therefore
statically guarantee that only calls that will return a valid result are
executed.

By incorporating pointers to open connections and prepared statements into the
resource associated with the effect, we introduce a further layer of
abstraction, in order to hide implementation details from the developer and
encourage cleaner, less error-prone code.

@@

String ->
Eff IO [SQLITE ()] (Either String (List (String, String)))

@@

\begin{SaveVerbatim}{testsel}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either String (List (String, String)))
testSelect dept = do
  open_db <- openDB "people.db"
  if open_db then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if sql_prep_res then do 
      startBind
      bindText 1 dept
      finishBind
      beginExecution
      results <- collectResults
      finaliseStatement
      closeDB
      return $ Right results
    else do err <- stmtFail
            return $ Left err
  else do err <- connFail
          return $ Left err 
\end{SaveVerbatim}

@@

\texttt{CollectResults} makes a call to
\texttt{nextRow} in order to make the next row of the result set available for
processing, and uses \texttt{getColumnText} and \texttt{getColumnInt}
to retrieve the data from the database. This function is then
recursively called until there are no more rows to process.

\begin{SaveVerbatim}{collectresty}

collectResults : Eff IO 
            [SQLITE (SQLiteRes PreparedStatementExec)] 
            (List (String, Int))
\end{SaveVerbatim}

\begin{SaveVerbatim}{collectres}
collectResults = do
  step_result <- nextRow
  case step_result of
      StepComplete => do name <- getColumnText 1
                         age <- getColumnInt 2
                         xs <- collectResults
                         return $ (name, age) :: xs
      NoMoreRows => return []
      StepFail => return []

@@

\noindent
Using these, we may implement a function, \texttt{ExecuteSelect}, which, given
a query, a list of variables to bind and their associated indices within the
query, and a function which is used to extract information out of each returned
database row, returns either a \texttt{ResultSet} or an error.

@@

executeSelect : 
  String -> String -> List (Int, DBVal) -> 
  Eff m [SQLITE (SQLiteRes PreparedStatementExec)] 
                 (List DBVal) -> 
  Eff m [SQLITE ()] (Either String ResultSet)

>>>>

In \idris{}, types are first-class, meaning that they may be treated like other
terms in computations. We may therefore define the following type synonym, used within the CGI section of the framework to denote an effectual CGI program:

@@

SQLite\footnote{\texttt{http://www.sqlite.org}} is a lightweight SQL database
engine often used as simple, structured storage for larger applications. We
make use of SQLite to demonstrate a resource usage protocol for database access due to its simplicity, although we envisage that these
concepts would be applicable to more complex database management systems.

@@

Figure~\ref{fig:sqlitestates} shows a resource usage protocol for database
access, which we have implemented for the SQLite library. Although some additional states are used to capture failing computations, these are omitted from the diagram. The protocol is encapsulated
by the \texttt{Sqlite} effect (Figure \ref{fig:dbeffect}; we again omit some operations, such as those to
bind and retrieve data types other than \texttt{String}, in the interest of
brevity).

\begin{SaveVerbatim}{databaseeff}
{-                           { Input resource type }    { Output resource type }             { Value }           -}
data Sqlite : Effect where
  OpenDB            : DBName -> 
                      Sqlite ()                         (Either () SQLiteConnected)          (Either SQLiteCode ())
  CloseDB           : Sqlite (SQLiteConnected)          ()                                   ()
  PrepareStatement  : QueryString -> 
                      Sqlite (SQLiteConnected)          (Either (SQLitePSFail) 
                                                          (SQLitePSSuccess Binding))         (Either SQLiteCode ())
  BindInt           : ArgPos -> Int -> 
                      Sqlite (SQLitePSSuccess Binding)  (SQLitePSSuccess Binding)            ()
  FinishBind        : Sqlite (SQLitePSSuccess Binding)  (Either SQLiteFinishBindFail
                                                          (SQLitePSSuccess Bound))           (Maybe BindError)
  ExecuteStatement  : Sqlite (SQLitePSSuccess Bound)    (Either (SQLiteExecuting InvalidRow)
                                                          (SQLiteExecuting ValidRow)         StepResult
  RowStep           : Sqlite (SQLiteExecuting ValidRow) (Either (SQLiteExecuting InvalidRow)
                                                          (SQLiteExecuting ValidRow))        StepResult
  GetColumnText     : Column -> 
                      Sqlite (SQLiteExecuting ValidRow) (SQLiteExecuting ValidRow)           String

  CleanupPSFail     : Sqlite (SQLitePSFail)             ()                                   ()
  CleanupBindFail   : Sqlite (SQLiteFinishBindFail)     ()                                   ()

@@

%TODO: Fix up the labels
\begin{figure}
\centering
\scalebox{0.8}{
\begin{tikzpicture}[>=latex]
  \tikzset{every loop/.style={min distance=10mm,looseness=5}}
  \tikzstyle{state} = [draw, very thick, fill=white, rectangle, minimum height=3em, minimum width=7em, align=center, node distance=7em, font={\sffamily\bfseries}]

@@

  \node[initial,state] (A)              {Uninitialised};
  \node[state]         (B) [below of=A] {SQLiteConnected};
  \node[state]         (C) [below of=B] {SQLitePSSuccess Binding};
  \node[state]         (D) [below of=C] {SQLitePSSuccess Bound};
  \node[state]         (E) [below left of=D, xshift=-4em] {SQLiteExecuting \\ ValidRow};
  \node[state]         (F) [below right of=D, xshift=4em] {SQLiteExecuting \\ InvalidRow};

  \path (A) edge[stateEdge]   node[edgeLabel, xshift=2.5em] {\texttt{openDB}} (B)
        (B) edge[stateEdge, bend right=80]   node[edgeLabel, text width=2cm, xshift=4.5em] {\texttt{closeDB, connFail}} (A)
        (B) edge[stateEdge]   node[edgeLabel, xshift=5em] {\texttt{prepareStatement}} (C)
        (C) edge[stateEdge]   node[edgeLabel, xshift=3.25em, yshift=-0.75em] {\texttt{finishBind}} (D)
        (D) edge[stateEdge]   node[edgeLabel, text width=2cm, xshift=-4.5em, yshift=-1em] {\texttt{executeStatement, nextRow}} (F)
        (D) edge[stateEdge]   node[edgeLabel, xshift=3em] {} (E)
        (E) edge[stateEdge, bend left=35]   node[edgeLabel, xshift=3em, yshift=-5em] {\texttt{finaliseValid}} (A)
        (F) edge[stateEdge, bend right=35]   node[edgeLabel, xshift=-3em, yshift=-5em] {\texttt{finaliseInvalid}} (A)
        (E) edge[stateEdge]   node[edgeLabel, yshift=-1em] {\texttt{nextRow}} (F)
        (F) edge[stateEdge, bend left=35]   node[edgeLabel, yshift=-1em] {\texttt{reset}} (E)
        (E) edge[stateEdge, loop below]   node[edgeLabel] {\texttt{nextRow}} (E);

\end{tikzpicture}
}
\caption{Database Resource Usage Protocol}
\label{fig:sqlitestates}
\end{figure}

@@

\end{SaveVerbatim}
\useverb{sqliteexecuting}
\noindent
Column access functions may only be called when there is a valid row in the buffer, as signified by the input resource of each operation.

@@

\noindent
In order to provide the necessary static guarantee that there is a valid row to process in the buffer, we make use of the \texttt{if\_valid} construct. The \texttt{nextRow} function will either fetch a row into the buffer, or indicate that a row could not be loaded, as shown by its type:

\begin{SaveVerbatim}{nextrow}
nextRow : EffM IO [SQLITE (SQLiteExecuting ValidRow)] 
                  [SQLITE (Either (SQLiteExecuting InvalidRow)
                                  (SQLiteExecuting ValidRow))] 
                  StepResult
\end{SaveVerbatim}
\useverb{nextrow}

\noindent
The \texttt{if\_valid} construct provides failure checking functionality, allowing different operations to be performed depending on whether or not a row was successfully fetched for processing.
%If a failure happens at any point during the computation, the resource is
%updated to reflect the failure: if, for example, the library failed to create a
%connection to the database, the resource value would be updated to
%\texttt{OpenConn InvalidConn}. At this point, no further side-effecting
%requests are made to the underlying SQLite library, in order to ensure safety.
%The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and
%\texttt{executeFail} utility functions allow for failures, once detected, to be
%handled by executing the appropriate sequence of state transition functions to
%dispose of any open resources and return to the initial protocol state.

@@

By incorporating pointers to open connections and prepared statements into the
resource associated with the effect, we introduce a further layer of
abstraction, which hides implementation details from the developer and
encourages less error-prone code.

@@

String ->
Eff IO [SQLITE ()] (Either QueryError (List (String, String)))

@@

\begin{SaveVerbatim}{testsel}
testSelect : String -> Eff IO [SQLITE ()] 
             (Either QueryError (List (String, String)))
testSelect dept = do
  db_res <- openDB "people.db"
  if_valid then do
    let sql = "SELECT name, address FROM `staff` 
                    WHERE dept = ?;"
    sql_prep_res <- prepareStatement sql
    if_valid then do 
      bindText 1 dept
      bind_res <- finishBind
      if_valid then do
        executeStatement
        results <- collectResults
        finaliseInvalid
        closeDB
        Effects.pure $ Right results
      else do
        cleanupBindFail
        Effects.pure $ Left (getBindError bind_res)
    else do
      cleanupPSFail
      Left $ getQueryError sql_prep_res
  else do 
    Effects.pure $ Left (getQueryError db_res)
\end{SaveVerbatim}

@@

\texttt{CollectResults} operates on the row currently held in the buffer. Firstly, the function checks, through the use of the \texttt{if\_valid} construct, that there is a valid row in the buffer, and if so uses the \texttt{getColumnText} and \texttt{getColumnInt} functions
to retrieve the data from the database. This function is then
called recursively until there are no more rows to process.

\begin{SaveVerbatim}{collectresty}
collectResults :
  EffM IO [SQLITE (Either (SQLiteExecuting InvalidRow) 
                          (SQLiteExecuting ValidRow))] 
          [SQLITE (SQLiteExecuting InvalidRow)] 
          (List (String, Int))
\end{SaveVerbatim}

\begin{SaveVerbatim}{collectres}
collectResults = 
  if_valid then do name <- getColumnText 1
                   age <- getColumnInt 2
                   step_res <- nextRow
                   xs <- collectResults
                   return $ (name, age) :: xs
  else return []

@@

\noindent
Using these, we may implement a function, \texttt{executeSelect}, which, given
a query, a list of variables to bind and their associated indices within the
query, and a function which is used to extract information out of each returned
database row, returns either a \texttt{ResultSet} or an error.

@@

executeSelect : 
  String -> String -> List (Int, DBVal) -> 
  (Eff IO [SQLITE (SQLiteExecuting ValidRow)] (List DBVal)) -> 
  Eff IO [SQLITE ()] (Either QueryError ResultSet)
